---
title: "Chapter 1 - Introduction"
author: Jeremy Buss
output: html_document
date: "`r Sys.Date()`"
categories:
  - John E. Freund's Mathematical Statistics with Applications - Miller/Miller
  - Statistics
draft: yes
katex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r message=FALSE, warning=FALSE, include=FALSE}
install.packages("blogdown")
```



## 1.2 Combinatorial Methods

**Theorem 1.1** If an operation consists of two steps, of which the first can be done in \(n_1\) ways and for each of these the second can be done in \(n_2\) ways, then the whole operation can be done in \(n_1 \cdot n_2\) ways.

**Theorem 1.2** If an operation consists of \(k\) steps, of which the first can be done in \(n_1\) ways, for each of these the second step can be done in \(n_2\) ways, for each of the first two the third step can be done in \(n_3\) ways, and so forth, then the whole operation can be done in \(n_1 \cdot n_2 \cdot \ldots \cdot n_k\) ways.

**Permutation** - a distinct arrangement of \(n\) different elements of a set

**Theorem 1.3** The number of permutations of \(n\) distinct objects is \(n!\)

**Theorem 1.4** The number of permutations of \(n\) distinct objects taken \(r\) at a time is
\[{}_nP_r = \frac{n!}{(n-r)!}\]
for \(r=0,1,2,\ldots,n\).

**Proof** The formula \({}_nP_r=n(n-1) \cdot \ldots \cdot (n-r+1)\) cannot be used for \(r=0\), but we do have

\[{}_nP_0 = \frac{n!}{(n-0)!}=1\]

For \(r=1,2,\ldots,n,\) we have

\[\begin{align}
{}_nP_r&=n(n-1)(n-2) \cdot \ldots \cdot (n-r-1)\\
&=\frac{n(n-1)(n-2) \cdot \ldots \cdot (n-r-1)(n-r)!}{(n-r)!}\\
&=\frac{n!}{(n-r)!}
\end{align}
\]

**circular permutation** - permutation for which objects are arranged in a circle

**Theorem 1.5** The number of permutations of \(n\) distinct objects arranged in a circle is \((n-1)!\)

**Theorem 1.6** The number of permutations of \(n\) objects of which \(n_1\) are of one kind, \(n_2\) are of a second kind, \(\ldots, n_k\) are of a \(k\)th kind, and \(n_1+n_2+\dots+n_k=n\) is

\[\frac{n!}{n_1! \cdot n_2! \cdot \ldots \cdot n_k!}\]

**Combination** - is a selection of \(r\) objects taken from \(n\) distinct objects without regard to the order of selection

**Theorem 1.7** The number of combinations of \(n\) distinct objects taken \(r\) at a time is 

\[\dbinom n r = \frac{n!}{r!(n-r)!}\]

for \(r=0,1,2,\ldots,n\)

**Partition (words)** - separate \(n\) distinct objects into \(k\) subsets, which requires that each of the \(n\) objects must belong to one and only one of the subsets

**Partition (symbolically)** - the subsets \(A_1, A_2, \ldots ,A_k\) constitute a partition of set \(A\) if \(A_1 \cup A_2 \cup \ldots \cup A_k = A\) and \(A_i \cap A_j = \emptyset\) for all \(i \ne j\)

**Theorem 1.8** The number of ways in which a set of \(n\) distinct objects can be partitioned into \(k\) subsets with \(n_1\) objects in the first subset, \(n_2\) objects in the second subset, \(\ldots\), and \(n_k\) objects in the \(k\)th subset is 

\[ \dbinom {n} {n_1, n_2, \ldots, n_k}=\frac{n!}{n_1! \cdot n_2! \cdot \ldots \cdot n_k!} \]

**Proof** Since the \(n_1\) objects going into the first subset can be chosen in \(\dbinom n {n_1} \) ways, the \(n_2\) objects going into the second subset can then be chose in \( \dbinom {n-n_1} {n_2}\) ways, the \(n_3\) objects going into the third subset can then be chosen in \( \dbinom {n-n_1-n_2} {n_3}\) ways, and so forth, it follows by Theorem 1.2 that the total number of partitions is

\[ 
\begin{align}
\dbinom {n} {n_1,n_2,\ldots,n_k} &= \dbinom n {n_1} \cdot \dbinom {n - n_1} {n_2} \cdot \ldots \cdot \dbinom {n-n_1-n_2-\cdots -n_{k-1}}{n_k}\\
&= \frac{n!}{n_1! \cdot (n-n_1)!} \cdot \frac{(n-n_1)!}{n_2! \cdot (n-n_1-n_2)!} \cdot \ldots \cdot \frac{(n-n_1-n_2-\cdots-n_{k-1})!}{n_k! \cdot 0!}\\
&= \frac{n!}{n_1! \cdot n_2! \cdot \ldots \cdot n_k!}
\end{align}
\]


## 1.3 Binomial Coefficients

**Theorem 1.9** \((x+y)^n=\sum^n_{r=0} \dbinom n r x^{n-r} y ^r\) for any positive integer \(n\)

**Binomial Coefficients** the coefficient of  \(x^{n-r} y ^r\) in the binomial expansion of \((x+y)^n\)

**Theorem 1.10** For any positive integers \(n\) and \(r=0,1,2,\ldots,n,\)

\[\dbinom n r = \dbinom n {n-r}\]

**Proof** We might argue that when we select a subset of \(r\) objects from a set of \(n\) distinct objects, we leave a subset of \(n-r\) objects; hence, there are as many ways of selecting \(r\) objects as there are ways of leaving (or selecting) \(n-r\) objects. To prove the theorem algebraically, we write

\[
\begin{align}
\dbinom n {n-r} &= \frac{n!}{(n-r)![n-(n-r)]!} = \frac{n!}{(n-r)!r!}\\
&= \frac{n!}{r!(n-r)!} = \dbinom n r
\end{align}
\]

**Theorem 1.11** For any positive integer \(n\) and \(r=1,2,\ldots, n-1\),

\[\dbinom n r = \dbinom {n-1} r + \dbinom {n-1} {r-1}\]

**Proof** Substituting \(x=1\) into \((x+y)^n\), let us write \((1+y)^n = (1+y)(1+y)^{n-1}=(1+y)^{n-1}+y(1+y)^{n-1}\) and equate the coefficient of \(y^r\) in \((1+y)^{n}\) with that in \((1+y)^{n-1}+y(1+y)^{n-1}\). Since the coefficient of \(y^r\) in \((1+y)^{n}\) is \(\dbinom n r \) and the coefficient of \(y^r\) in \((1+y)^{n-1}+(1+y)^{n-1}\) is the sum of the coefficient of \(y^r\) in \((1+y)^{n-1}\), that is, \(\dbinom {n-1} r\), and the coefficient of \(y^{r-1}\) in \((1+y)^{n-1}\), that is, \(\dbinom {n-1} {r-1} \), we obtain 

\[\dbinom n r = \dbinom {n-1} r + \dbinom {n-1} {r-1}\]

which completes the proof.

**Pascal's Triangle (words)** The first and last entries of each row are the numeral "1" each other entry in any given row is obtained by adding the two entries in the preceding row immediately to its left and its right

**Pascal's Triangle (numerical)**
\[
\begin{align}
& \, \quad \quad 1\\
& \; \quad 1 \quad 1 \\
& \; 1 \quad 2 \quad 1\\
1 & \quad 3 \quad 3 \quad 1 \\
1 \quad& 4 \quad 6 \quad 4 \quad 1 \\
....&.............
\end{align}
\]

The subset can't contain more elements than the original set; i.e. \(\dbinom n r = 0\) whenever \(n\) is a positive integer and \(r\) is a positive integer greater than \(n\). 

**Theorem 1.12** 

\[
\sum^k_{r=0} \dbinom m k \dbinom n {k-r} = \dbinom {m+n} k 
\]

**Proof** Using the same technique as in the proof of Theorem 1.11, let us prove this theorem by equating the coefficients of \(y^k\) in the expressions on both sides of the equation

\[(1+y)^{m+n}=(1+y)^m(1+y)^n\]

The coefficient of \(y^k\) in \((1+y)^{m+n}\) is \(\dbinom {m+n} k\), and the coefficient of \(y^k\) in 

\[(1+y)^m(1+y)^n = [ \dbinom m 0 + \dbinom m 1 y + \cdots + \dbinom m m y^m ] \times [ \dbinom n 0 + \dbinom n 1 y + \cdots + \dbinom n n y^n ] \]

is the sum of the prodcts that we obtain by multiplying the constant term of the first factor by the coefficient of \(y^k\) in the second factor, the coefficient of \(y\) in the first factor by the coefficient of \(y^{k-1}\) in the second factor, ..., and the coefficient of \(y^k\) in the first factor by the constant term of the second factor. Thus, the coefficient of \(y^k\) in \((1+y)^m(1+y)^n\) is

\[
\dbinom m 0 \dbinom n k + \dbinom m 1 \dbinom n {k-1} + \dbinom m 2 \dbinom n {n-2} + \cdots + \dbinom m k \dbinom n 0 = \sum^k_{r=0} \dbinom m r \dbinom n {k-r}
\]

and this completes the proof.

**Multinomial Coefficients** coefficients that arise in the expansion of \( (x_1+x_2+\cdots+x_k)^n \). The multinomial coefficient of the term \(x_1^{r_1} \cdot x_2^{r_2} \cdot \ldots \cdot x_k^{r_k} \) in the expansion of \( (x_1+x_2+\cdots+x_k)^n \) is 

\[
\dbinom n {r_1,r_2,\ldots,r_k} = \frac {n!}{r_1! \cdot r_2! \cdot \ldots \cdot r_k!}
\]