---
title: "Chapter 1 - Introduction"
author: Jeremy Buss
output: html_document
date: '2022-06-04'
categories:
  - Statistics
draft: yes
katex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r message=TRUE, warning=TRUE}
install.packages("blogdown")
```


## 1.2 Combinatorial Methods

**Theorem 1.1** If an operation consists of two steps, of which the first can be done in \(n_1\) ways and for each of these the second can be done in \(n_2\) ways, then the whole operation can be done in \(n_1 \cdot n_2\) ways.

**Theorem 1.2** If an operation consists of \(k\) steps, of which the first can be done in \(n_1\) ways, for each of these the second step can be done in \(n_2\) ways, for each of the first two the third step can be done in \(n_3\) ways, and so forth, then the whole operation can be done in \(n_1 \cdot n_2 \cdot \ldots \cdot n_k\) ways.

**Permutations** - a distinct arrangement of \(n\) different elements of a set

**Theorem 1.3** The number of permutations of \(n\) distinct objects is \(n!\)

**Theorem 1.4** The number of permutations of \(n\) distinct objects taken \(r\) at a time is
\[{}_nP_r = \frac{n!}{(n-r)!}\]
for \(r=0,1,2,\ldots,n\).
**Proof** The formula \({}_nP_r=n(n-1) \cdot \ldots \cdot (n-r+1)\) cannot be used for 

## 1.3 Binomial Coefficients

**Theorem 1.12** 

\[\sum^k_{r=0} \dbinom m k \dbinom n {k-r} = \dbinom {m+n} k \]

**Proof** Using the same technique as in the proof of Theorem 1.11, let us prove this theorem by equating the coefficients of \(y^k\) in the expressions on both sides of the equation

\[(1+y)^{m+n}=(1+y)^m(1+y)^n\]

The coefficient of \(y^k\) in \((1+y)^{m+n}\) is \(\dbinom {m+n} k\), and the coefficient of \(y^k\) in 

\[(1+y)^m(1+y)^n = \]