---
title: "Chapter 5 - Special Probability Distributions"
author: Jeremy Buss
output: html_document
date: "`r Sys.Date()`"
categories:
  - John E. Freund's Mathematical Statistics with Applications - Miller/Miller
  - Statistics
draft: yes
katex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r message=FALSE, warning=FALSE, include=FALSE}
install.packages("blogdown")
```



## 5.1 Introduction

**parameters** - quantities that are constants for particular distributions, but can take on different values for different members of famiiles of distributions of the same kind

## 5.2 The Discrete Uniform Distribution

**Discrete Uniform Distribution** - A random variable \(X\) has a **discrete uniform distribution** and it is referred to as a discrete uniform random variable if and only if it probability distribution is given by

\[
f(x) = \frac{1}{k} \; \text{for} \, x = x_1, x_2, \dots, x_k
\]

where \(x_i \ne x_j\) when \(i \ne j\)

## 5.3 The Bernoulli Distribution

**Bernoulli Distribution** - A random variable \(X\) has a **Bernoulli distribution** and it is referred to as a Bernoulli random variable if and only if its probability distribution is given by

\[
f(x; \theta) = \theta ^x (1-\theta)^{1-x} \; \text{for} \, x=0,1
\]

**Bernoulli Trial** - an experiment to which a Bernoulli distribution applies; an experiment that has two possible outcomes "success" and "failure"

## 5.4 The Binomial Distribution

**Binomial Distribution** - A random variable \(X\) has a **Binomial distribution** and it is referred to as a binomail random variable if and only if its probability distribution is given by

\[
b(x; n, \theta)= \dbinom n x \theta^x (1-\theta)^{n-x} \; \text{for} \, x=0,1,2,\dots,n
\]

**Theorem 5.1 [Binomial Distribution - Complementary]**

\[
b(x; n, \theta)=b(n-x; n, 1 - \theta)
\]

**Theorem 5.2 [Binomial Distribution - Mean and Variance]**
The mean and variance of the binomial distribution are

\[
\mu = n\theta \;\; \text{and} \;\; \sigma^2 = n \theta(1-\theta)
\]

**Theorem 5.2 [Binomial Distribution - Mean and Variance - Proof]**

\[
\]

where we omitted the term corresponding to \(x=0\), which is 0, and canceled the \(x\) against the first factor of \(x!=x(x-1)\) in the denominator of \(\dbinom n x\). Then, factoring out the factor \(n\) in \(n! = n(n-1)!\) and one factor \(\theta\), we get

\[
\]

and, letting \(y=x-1\) and \(m = n-1\), this becomes

\[
\]

since the last summation is the sum of all the values of a binomial distribution with the parameters \(m\) and \(\theta\), and hence equal to 1.

To find expressions for 

## 5.5 The Negative Binomial and Geometric Distributions

## 5.6 The Hypergeometric Distribution

## 5.7 The Poisson Distribution

## 5.8 The Multinomial Distribution

## 5.9 The Multivariate Hypergeometric Distribution


